Nonparametric Clustering from Energy Statistics

### 2017-06-21 qcqp_clustering/draft/energy_clustering.tex
working on draft

%\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
\documentclass{article}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2017}

\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[inline]{enumitem}

%\usepackage{cite}
\usepackage{bbm}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

@@

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\delta_{#1}}

\title{Nonparametric Clustering Based on Energy Statistics}

\author{
Guilherme Fran\c ca \\
Johns Hopkins University\\
\texttt{guifranca@gmail.com} \\
%% examples of more authors
\And
Joshua T. Vogelstein \\
Johns Hopkins University\\
\texttt{jovo@jhu.edu}
}

%\author{Guilherme Fran\c ca}
%\email{guifranca@gmail.com}
%\affiliation{Johns Hopkins University, Center for Imaging Science}

%\author{Joshua Vogelstein}
%\email{jovo@jhu.edu}
%\affiliation{Johns Hopkins University, Center for Imaging Science}

@@

@@

\begin{abstract}
blabla
\end{abstract}

@@

In this section we briefly review the main concepts from energy
statistics and its relation to reproducing kernel Hilbert spaces 
(RKHS), which form the basis of our work.
For more details we refer the reader
to 
\cite{Szkely2013} 
(and references therein) and also \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}\label{eq:energy}
\Energy(P, Q) = 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}\label{eq:energy2}
\Energy_\alpha(P, Q) = 2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha},
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have that
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$, showing that
equality to zero only requires
equality of the means, and thus it does not imply equality of distributions.

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$ and replace the Euclidean norm by
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, i.e.
$\| X - Y\| \to \rho(X,Y)$, where $\rho$ is a so-called \emph{semimetric
of negative type},
which satisfy 
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n \alpha_i \alpha_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$, and $\alpha_i \in \mathbb{R}$ such that
$\sum_{i=1}^n \alpha_i = 0$.
In this case, there is a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. 
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for any $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
any $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ there is a unique function
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS.
From this we have $\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$. 
This implies
that $\kk(x,y)$ is symmetric and positive semi-definite, 
$\sum_{i,j=1}^n c_i c_j
\kk(x_i,x_j) \ge 0$ for $c_i,c_j \in \mathbb{R}$.

The Moore-Aronszajn theorem establishes the converse \cite{Aronszajn}.
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical feature map. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively $\kkk_P = \int \kk(\cdot, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$. This is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}\label{eq:mmd}
\gamma_\kk(P,Q) = \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Therefore, in practice, we can estimate the inner product between the 
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive semi-definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
be a semimetric, 
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
\end{equation}
Then, $\kk$ is positive semi-definite if and only if $\rho$ is of negative type
\eqref{eq:negative_type}.
Here we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\rho(x,y) = \kk(x,x) + \kk(y,y) -2\kk(x,y) = \| \kkk_x - \kkk_y
\|^2_{\Hk},
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
We say that the kernel $\kk$ generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$, they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
\begin{align}
\Energy(P, Q) &\equiv 2\E \, \rho(X,Y) - \E \, \rho(X,X') - \E \, \rho(Y,Y') 
\label{eq:EphoDef}\\
&= 2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] \\
&=2 \gamma_\kk^2(P,Q) .
\end{align}
This result follows simply by substituting \eqref{eq:gen_kernel} into
\eqref{eq:EphoDef}, and using \eqref{eq:mmd2}.
Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
can compute the energy distance using the inner product of $\Hk$. Moreover,
this can be computed from the data according to \eqref{eq:inner_data}.

Finally, let us recall the main formulas for test statistics
\cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,x_2,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$ is in some space of 
negative type \eqref{eq:negative_type},
and a partition $\mathbb{X} = \cup_{j=1}^k \C_j$ where
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:EphoDef}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  j < l \le k } \dfrac{n_j n_{l}}{2 n} \left[
2 g(\C_j, \C_l) - 
g(\C_j, \C_j) - 
g(\C_l, \C_l)
\right].
\end{equation}
Given a set of distributions
$\{ P_j\}_{j=1}^k$ where $x \in \C_j \sim P_j$, the quantity
$S$ provides
a \emph{nonparametric} test statistic for equality of distributions.
Under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$, $S$ is small, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
$S \to \infty$ as the sample size is large, $n \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

@@

This section contains the main results of this paper, where 
we propose an optimization problem for clustering data
based on energy statistics and RKHS introduced in the previous section.

Based on the test statistic \eqref{eq:between}, the obvious
criterion for clustering data is to 
maximize $S$, which intuitivelly makes the distribution of each
cluster $\C_j$ as different
as possible from the other ones. However, the following 
simple result
shows that this is equivalent to minimizing
\eqref{eq:within}, which has a more convenient form.

\begin{proposition}
\label{th:minimize}
Let $\mathbb{X} = \{x_i\}_{i=1}^{n}$ where 
$x_i$ lives in a space $\mathcal{X}$ endowed with a semi-metric $\rho$ of
negative type. For a fixed integer $k$,
the partition
$\mathbb{X} = \cup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes $S$ if and only if
\begin{equation}
\label{eq:minimize}
\min_{\{ \C_j \} } W(\{ 
\C_1, \dotsc, \C_k
\}).
\end{equation}
\end{proposition}
\begin{proof}
It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
\begin{equation}
W + S = \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}). 
\end{equation}
Note, however, that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent on the choice of partition. Therefore, maximizing
$S$ is equivalent to minimizing~$W$.
\end{proof}

Thus, given $k$, our clustering problem ammounts to
finding the best partition of the data through solving~\eqref{eq:minimize}.
In this setting, each datapoint belongs to one and only one cluster
(hard assignments).

Based on 
\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, assume that 
the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$. 
Let us define  the kernel matrix
$G \in \mathbb{R}^{n\times n}$ such that
\begin{equation}
\label{eq:kernel_matrix}
G_{ij} \equiv \kk(x_i, x_j).
\end{equation}
Let $Z$ be the label matrix, $Z \in \{ 0,1 \}^{n\times k}$, 
with only one nonvanishing entry per row indicating to which cluster (columns)
each point (rows) belongs to. This matrix satisfy
$Z^\top Z = D$ where $D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}

Our next result reveals the optimization problem behind \eqref{eq:minimize},
which is in general NP-hard since
it is a quadratically constrained quadratic problem (QCQP).

>>>>

\documentclass[aps,preprint,nofootinbib,floatfix]{revtex4-1}
%\documentclass{article}

@@

% to compile a camera-ready version, add the [final] option, e.g.:
%\usepackage[final]{nips_2017}

%\usepackage{nicefrac}       % compact symbols for 1/2, etc.
%\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm,amscd,bm,bbm}
%\usepackage{mathtools}
%\usepackage{printlen}
%\usepackage[inline]{enumitem}
%\usepackage{cite}
%\usepackage{bbm}

@@

\usepackage{algorithmic}
\usepackage{algorithm}

@@

\newcommand\Energy{\mathcal{E}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\OO{{\mathcal{O}}}
%\newcommand\Zt{\widetilde{Z}}
\newcommand\Zt{Y}
%\newcommand{\Ind}[1]{\delta_{#1}}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}

@@

\title{Nonparametric Clustering from  Energy Statistics}

\author{Guilherme Fran\c ca}
\email{guifranca@gmail.com} 
%\affiliation{Johns Hopkins University}

\author{Joshua T. Vogelstein}
\email{jovo@jhu.edu}

\affiliation{Johns Hopkins University}

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's,
inspired by Newton's potential energy between massive bodies. The idea
is to associate a statistical potential energy to observations such that 
minimum energy is achieved under the null hypothesis of equality of 
distributions. This 
was further generalized for probability 
distributions on arbitrary metric spaces,
and more recently, a connection with kernels in RKHS was established.
Nevertheless, although extensively used by the statistics community, it was
not incorporated in machine learning problems.
In this paper, we consider the problem of clustering data based
energy statistics theory. 
We provide a precise mathematical formulation, obtaining
a quadratically constrained optimization problem (QCQP). We show that
this optimization problem is equivalent to kernel $k$-means
optimization problem, however,
energy statistics is able to fix the kernel choice. 
This method is nonparametric, and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an algorithm
to find local optimizers of this QCQP problem based on the energy cost
of moving points to different clusters. We then compare this algorithm with
kernel $k$-means, standard $k$-means, and GMM. We test our method
on synthetic and real data experiments.
Our results show that energy statistics based clustering outperforms
these most used clustering algorithms.
\end{abstract}

@@

@@

In this section we briefly review the main concepts from energy
statistics and its relation to reproducing kernel Hilbert spaces 
(RKHS) which form the basis of our work.
For more details we refer the reader
to \cite{Szkely2013} (and references therein) and 
also \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions and
$\Energy^{1/2}$ is
a metric on the space of distributions.

The energy distance can be generalized as, for instance,
\begin{equation}\label{eq:energy2}
\Energy_\alpha(P, Q) \equiv 
2 \E \| X - Y\|^{\alpha} - \E \| X - X' \|^{\alpha} - 
\E \| Y - Y' \|^{\alpha},
\end{equation}
where $0<\alpha\le 2$. This quantity is also nonnegative,
$\Energy_\alpha(P,Q) \ge 0$. Furthermore, for $0<\alpha<2$ we have 
$\Energy_\alpha(P,Q) = 0$ if and only if $P=Q$, while for $\alpha=2$ 
we have $\Energy_2(P,Q) = 2\| \E(X) - \E(Y) \|^2$ showing that
equality to zero only requires
equality of the means and thus $\Energy_2(P,Q)=0$ does 
not imply equality of distributions.

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is a space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which satisfies
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n \alpha_i \alpha_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $\alpha_i \in \mathbb{R}$ such that
$\sum_{i=1}^n \alpha_i = 0$. Then $\mathcal{X}$ is called a space of
negative type.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy} obtaining
the energy distance as
\begin{equation}\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. 
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. We first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$ 
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
this inner product is real,
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
\langle \kkk_x, \kkk_y \rangle$, we immediately have that
the kernel is symmetric,
$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semi-definite, $\bm{w}^\top G \bm{w} \ge 0$ for any vector
$\bm{w} \in \mathbb{R}^n$.

The Moore-Aronszajn theorem establishes the converse \cite{Aronszajn}.
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical feature map. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P = \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$. This is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Therefore, in practice, we can estimate the inner product between the 
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive semidefinite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) = \tfrac{1}{2} \left\{  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right\}.
\end{equation}
Thus, $\kk$ is positive semidefinite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
Here we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
We say that the kernel $\kk$ generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3} and using \eqref{eq:mmd2} yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Since $\gamma_k^2(P, Q) = \| \kkk_P - \kkk_Q \|^2_{\Hk}$ we
can compute the energy distance using the inner product of $\Hk$. Moreover,
this can be computed from the data according to \eqref{eq:inner_data}.

Finally, let us recall the main formulas for test statistics
of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right].
\end{equation}
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric} test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$. Next we show how we can use $S$ for clustering
data.

@@

This section contains the main results of this paper, where 
we formulate an optimization problem for clustering 
based on energy statistics and RKHS introduced in the previous section.

Due to the test statistic \eqref{eq:between} for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$, which intuitively makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, $S$ should attain a maximum when the points are correctly
classified since $S$ measures a cost between different clusters.
However, the following 
straightforward result
shows that maximizing \eqref{eq:between} is equivalent to minimizing
\eqref{eq:within}, which has a more convenient form.

\begin{proposition}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type \eqref{eq:negative_type}. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes \eqref{eq:between} if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where $W$ is given by \eqref{eq:within}.
\end{proposition}
\begin{proof}
It can be shown that the total dispersion of the data obeys \cite{Szkely2013}
\begin{equation}
W + S = \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}). 
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ is equivalent to minimizing~$W$.
\end{proof}

Thus, for a given $k$, the clustering problem amounts to
finding the best partition of the data by solving~\eqref{eq:minimize}.
Notice that this is a hard assignment clustering problem.

Now we show how to formulate problem \eqref{eq:minimize} in the corresponding
RKHS.
Based on 
\eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, assume that 
the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$. 
Let us define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
\vdots & \vdots & \ddots  & \vdots \\
\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfy
$Z^\top Z = D$ where $D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.

Our next result reveals the optimization problem behind \eqref{eq:minimize},
which is NP-hard since
it is a quadratically constrained quadratic problem (QCQP).
